---
title: "Geopolitical Dashboard"
format: 
  dashboard: 
    theme: flatly
    highlight-style: github
    mainfont: "Inter"
    orientation: columns
nav-buttons: [github]
github: https://github.com/duck217/project-1
logo: globe-1.png
logo-alt: "Blue globe logo"
server: shiny
---

# GDP Data Across Regimes

```{r}
#Chunk used to load needed packages and data sets
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(tidyverse)
  library(countrycode)
  library(glue)
  library(jsonlite)
  library(palmerpenguins)
  library(gt)
  library(readr)
  library(tidyr)
  library(purrr)
  library(ggridges)
  library(stringr)
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(gganimate)
  library(plotly)
  library(tigris)
  library(leaflet)
  library(shiny)
  library(bslib)
})

meta_data <- suppressMessages(read_csv("meta_data.csv", show_col_types = FALSE))
```

```{r}
#| message: false
#| warning: false

combined <- meta_data |>
  filter(!is.na(gdp_per_capita_nominal), regime_category != "NA") |>
  group_by(regime_category) |>
  mutate(regime_group = if_else(n() < 100, "Territories & Colonies",
                                 first(regime_category))) |>
  ungroup()

mean_gdp <- combined |>
  group_by(year, regime_group) |>
  summarize(mean_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE),
            .groups = "drop")

dem_gdp <- combined |>
  mutate(democracy = if_else(is_democracy == 1, "Democracy", "Non-Democracy")) |>
  group_by(year, democracy) |>
  summarize(mean_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE),
            .groups = "drop")

```

## Plots 

### Mean GDP per Capita by Regime Type
::: {.card title = "Mean GDP per Capita by Regime Type"}

```{r}
#| label: fig-gdp-area
#| fig-alt: "Faceted area plot showing mean GDP per capita from 2000 to 2020 for seven regime types."

ggplot(data = mean_gdp,
       mapping = aes(x = year,
                     y = mean_gdp)) +
  geom_area(mapping = aes(fill = regime_group),
            alpha = 0.7,
            identity = "stack") +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "", y = "Mean GDP per Capita (USD)",
       title = "Mean GDP per Capita by Regime Type",
       subtitle = "Averaged across countries, 2000–2020",
       fill = "Regime Type") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())
```
:::

### GDP per Capita: Democracies vs Non-Democracies 
::: {.card title = "Mean GDP per Capita: Democracies vs Non-Democracies"}

```{r}
#| label: fig-gdp-demo-animated
#| fig-alt: "Animated line plot comparing mean GDP per capita between democracies and non-democracies from 2000 to 2020."

p <- ggplot(dem_gdp,
            aes(x = year, y = mean_gdp,
                color = democracy,
                group = democracy)) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = comma) +
  labs(x = "",
       y = "Mean GDP per Capita (USD)",
       title = "Democracy vs Non-Democracy Countries",
       color = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) +
  transition_reveal(year)

animate(p)
```
:::


## Tables 
::: panel-tabset

### GDP per Capita by Regime Type

```{r}
#| label: tbl-gdp-regime
combined |>
  group_by(regime_group) |>
  summarize(
    mean_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE),
    median_gdp = median(gdp_per_capita_nominal, na.rm = TRUE),
    n = n()
  ) |>
  arrange(desc(mean_gdp)) |>
  gt() |>
  tab_header(title = "GDP per Capita by Regime Type") |>
  cols_label(regime_group = "Regime Type",
             mean_gdp = "Mean",
             median_gdp = "Median",
             n = "Observations") |>
  tab_spanner(label = "GDP per Capita (USD)",
              columns = c(mean_gdp, median_gdp)) |>
  fmt_currency(columns = c(mean_gdp, median_gdp),
               currency = "USD", decimals = 2) |>
  fmt_number(columns = n, decimals = 0) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  data_color(columns = c(mean_gdp, median_gdp),
             palette = c("#deebf7", "#08519c")) |>
  tab_caption("GDP per capita sorted in descending mean GDP by regime type")
```


### Mean GDP per Capita: Democracies vs Non-Democracies

```{r}
#| label: tbl-gdp-democracy
dem_gdp |>
  pivot_wider(names_from = democracy, values_from = mean_gdp) |>
  gt() |>
  tab_header(title = "Mean GDP per Capita: Democracies vs Non-Democracies") |>
  fmt_currency(columns = c(Democracy, `Non-Democracy`), decimals = 0) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_caption("Mean GDP per capita compared by democracy status, 2000–2020.")
```


:::

# Diving into USA 

```{r}
#| context: setup
#| message: false
#| warning: false

library(shiny)
library(readr)
library(dplyr)
library(stringr)
library(tigris)
library(leaflet)
library(scales)
library(sf)
library(gt)

SAINC1_ALL_AREAS_1929_2025 <- read_csv(
  "SAINC1__ALL_AREAS_1929_2025.csv",
  show_col_types = FALSE
)

us_data <- SAINC1_ALL_AREAS_1929_2025 |>
  filter(Description == "Per capita personal income (dollars) 2/") |>
  select(GeoName, "2025") |>
  rename(income_2025 = `2025`) |>
  mutate(
    GeoName = str_remove(GeoName, " \\*$"),
    income_2025 = parse_number(as.character(income_2025))
  ) |>
  filter(GeoName %in% state.name)

states_sf <- states(
  cb = TRUE,
  class = "sf",
  progress_bar = FALSE
)

map_data <- states_sf |>
  filter(NAME %in% state.name) |>
  left_join(us_data, by = c("NAME" = "GeoName"))

colors <- colorNumeric(
  palette = c("#f7fbff", "#6baed6", "#08306b"),
  domain = map_data$income_2025,
  na.color = "transparent"
)
```

## {.sidebar}

```{r}
sliderInput(
  inputId = "income_min",
  label = "Minimum Income Per Capita",
  min = 0,
  max = ceiling(max(map_data$income_2025, na.rm = TRUE)),
  value = 0,
  step = 500,
  sep = ",",
  pre = "$"
)
```

## Column {width=35%}

### Why the United States?

::: {.card}

As a **presidential democracy**, the United States ranks among the highest GDP per capita countries in the dataset. Looking at state-level income shows whether economic output is **evenly spread** or **concentrated** in specific regions.

:::

### Filtered Income Table

::: {.card}

```{r}
gt_output("income_table")
```

:::

## Column {width=65%}

### Income per Capita Map

::: {.card}

```{r}
leafletOutput("us_map", height = 780)
```

:::

```{r}
#| context: server

filtered_states <- reactive({
  threshold <- input$income_min
  if (is.null(threshold)) threshold <- 0

  map_data |>
    filter(
      !is.na(income_2025),
      income_2025 >= threshold
    )
})

output$us_map <- renderLeaflet({
  leaflet(filtered_states(), options = leafletOptions(
    zoomControl = TRUE,
    attributionControl = FALSE
  )) |>
    addProviderTiles(providers$CartoDB.PositronNoLabels) |>
    setView(lng = -98.5, lat = 39.8, zoom = 4) |>
    addPolygons(
      fillColor = ~colors(income_2025),
      fillOpacity = 0.9,
      color = "white",
      weight = 1.4,
      label = ~paste0(NAME, ": $", comma(income_2025)),
      highlightOptions = highlightOptions(
        weight = 3,
        color = "black",
        bringToFront = TRUE
      )
    ) |>
    addLegend(
      position = "bottomright",
      pal = colors,
      values = na.omit(map_data$income_2025),
      title = "Income per capita",
      labFormat = labelFormat(prefix = "$", big.mark = ","),
      opacity = 1
    )
})

output$income_table <- render_gt({
  threshold <- input$income_min
  if (is.null(threshold)) threshold <- 0

  filtered_states() |>
    st_drop_geometry() |>
    transmute(
      State = NAME,
      income_2025 = income_2025
    ) |>
    arrange(desc(income_2025)) |>
    gt() |>
    tab_header(title = "Income per Capita by State") |>
    cols_label(
      State = "State",
      income_2025 = "Income per Capita"
    ) |>
    fmt_currency(
      columns = income_2025,
      currency = "USD",
      decimals = 0
    ) |>
    data_color(
      columns = income_2025,
      palette = c("#f7fbff", "#6baed6", "#08306b")
    ) |>
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels()
    ) |>
    tab_caption(
      paste0(
        "Showing states with income per capita of at least ",
        dollar(threshold, accuracy = 1),
        "."
      )
    )
})
```