---
title: "Geopolitical Dashboard"
format: 
  dashboard: 
    theme: flatly
    highlight-style: github
    mainfont: "Inter"
    orientation: columns
embed-resources: true
nav-buttons: [github]
github: https://github.com/duck217/project-1
logo: globe-1.png
logo-alt: "Blue globe logo"
---

# GDP Data Across Regimes

```{r}
#Chunk used to load needed packages and data sets
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(tidyverse)
  library(countrycode)
  library(glue)
  library(jsonlite)
  library(palmerpenguins)
  library(gt)
  library(tidyr)
  library(purrr)
  library(ggridges)
  library(stringr)
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(plotly)
  library(tigris)
  library(leaflet)
})

meta_data <- suppressMessages(read_csv("meta_data.csv", show_col_types = FALSE))
```

```{r}
#| message: false
#| warning: false

combined <- meta_data |>
  filter(!is.na(gdp_per_capita_nominal), regime_category != "NA") |>
  group_by(regime_category) |>
  mutate(regime_group = if_else(n() < 100, "Territories & Colonies",
                                 first(regime_category))) |>
  ungroup()

mean_gdp <- combined |>
  group_by(year, regime_group) |>
  summarize(mean_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE),
            .groups = "drop")

dem_gdp <- combined |>
  mutate(democracy = if_else(is_democracy == 1, "Democracy", "Non-Democracy")) |>
  group_by(year, democracy) |>
  summarize(mean_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE),
            .groups = "drop")

```

## Plots 

### Mean GDP per Capita by Regime Type
::: {.card title = "Mean GDP per Capita by Regime Type"}

```{r}
#| label: fig-gdp-area
#| fig-alt: "Faceted area plot showing mean GDP per capita from 2000 to 2020 for seven regime types."

ggplot(data = mean_gdp,
       mapping = aes(x = year,
                     y = mean_gdp)) +
  geom_area(mapping = aes(fill = regime_group),
            alpha = 0.7) +
  facet_wrap(~regime_group) +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "", y = "Mean GDP per Capita (USD)",
       title = "Mean GDP per Capita by Regime Type",
       subtitle = "Averaged across countries, 2000–2020",
       fill = "Regime Type") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())
```
:::

### GDP per Capita: Democracies vs Non-Democracies 
::: {.card title = "Mean GDP per Capita: Democracies vs Non-Democracies"}

```{r}
#| label: fig-gdp-demo-interactive
#| fig-alt: "Interactive line plot comparing mean GDP per capita between democracies and non-democracies from 2000 to 2020."

library(plotly)

p <- ggplot(dem_gdp, aes(x = year, y = mean_gdp, color = democracy)) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = comma) +
  labs(x = "", y = "Mean GDP per Capita (USD)",
       title = "Democracy vs Non-Democracy Countries",
       color = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

ggplotly(p)
```
:::


## Tables 
::: panel-tabset

### GDP per Capita by Regime Type

```{r}
#| label: tbl-gdp-regime
combined |>
  group_by(regime_group) |>
  summarize(
    mean_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE),
    median_gdp = median(gdp_per_capita_nominal, na.rm = TRUE),
    n = n()
  ) |>
  arrange(desc(mean_gdp)) |>
  gt() |>
  tab_header(title = "GDP per Capita by Regime Type") |>
  cols_label(regime_group = "Regime Type",
             mean_gdp = "Mean",
             median_gdp = "Median",
             n = "Observations") |>
  tab_spanner(label = "GDP per Capita (USD)",
              columns = c(mean_gdp, median_gdp)) |>
  fmt_currency(columns = c(mean_gdp, median_gdp),
               currency = "USD", decimals = 2) |>
  fmt_number(columns = n, decimals = 0) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  data_color(columns = c(mean_gdp, median_gdp),
             palette = c("#deebf7", "#08519c")) |>
  tab_caption("GDP per capita sorted in descending mean GDP by regime type")
```


### Mean GDP per Capita: Democracies vs Non-Democracies

```{r}
#| label: tbl-gdp-democracy
dem_gdp |>
  pivot_wider(names_from = democracy, values_from = mean_gdp) |>
  gt() |>
  tab_header(title = "Mean GDP per Capita: Democracies vs Non-Democracies") |>
  fmt_currency(columns = c(Democracy, `Non-Democracy`), decimals = 0) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_caption("Mean GDP per capita compared by democracy status, 2000–2020.")
```


:::

# Diving into USA 

```{r}
#| message: false
#| warning: false
library(readr)

SAINC1_ALL_AREAS_1929_2025 <- read_csv("SAINC1__ALL_AREAS_1929_2025.csv")

us_data <- SAINC1_ALL_AREAS_1929_2025

us_data <- us_data |>
  filter(Description == "Per capita personal income (dollars) 2/") |>
  select(GeoFIPS, GeoName, Region, "2025") |>
  rename(income_2025 = `2025`)

write.csv(us_data, file = "us_data.csv")
```

## Tables

### Why the United States?

As a **presidential democracy** country, the United States consistently ranks among the highest GDP per capita countries in our dataset, serving as an outlier. So, by examining how annual income is distributed across states reveals whether this economic output is **evenly spread** or **concentrated** in specific regions.

### Summary Table

```{r}
#| label: tbl-income-state

us_data |>
  mutate(
    GeoName = str_remove(GeoName, " \\*$"),
    income_2025 = as.numeric(income_2025)
  ) |>
  filter(
    GeoName != "United States",
    !GeoName %in% c("New England", "Mideast", "Great Lakes", "Plains",
                    "Southeast", "Southwest", "Rocky Mountain", "Far West")
  ) |>
  select(GeoName, income_2025) |>
  arrange(desc(income_2025)) |>
  gt() |>
  tab_header(title = "Income per Capita by State") |>
  cols_label(
    GeoName = "State",
    income_2025 = "Income per Capita"
  ) |>
  fmt_currency(columns = income_2025, currency = "USD", decimals = 0) |>
  data_color(
    columns = income_2025,
    palette = c("#f7fbff", "#6baed6", "#08306b")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_caption("Per capita personal income by U.S. state, sorted from highest to lowest.")
```

## Income per Capita by State

```{r}
#| label: fig-income-state
#| output: asis
#| fig-alt: "Map of the lower 48 U.S. states displaying income per capita by state using a color scale"

states_sf <- states(cb = TRUE, class = "sf",
                    progress_bar = FALSE)

map_data <- states_sf |>
  left_join(us_data, by = c("NAME" = "GeoName"))

colors <- scales::col_numeric(
  palette = c("#f7fbff", "#6baed6", "#08306b"),
  domain = map_data$income_2025)

leaflet(map_data, height = 500) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  addPolygons(fillColor = ~colors(income_2025),
              fillOpacity = 1,
              color = "white",
              weight = 1,
              label = ~paste0(NAME, ": $", scales::comma(income_2025)),
    highlightOptions = highlightOptions(weight = 2,
      color = "black",
      bringToFront = TRUE)) |>
  addLegend(position = "bottomright",
            pal = colors,
            values = ~na.omit(income_2025),
            title = "Income per capita",
            labFormat = labelFormat(prefix = "$",
                                    big.mark = ","),
            opacity = 1) |>
   setView(-96, 42, 4) 
```